\scuba: iterative map-making with the Sub-Millimetre User Reduction Facility

### 2012-08-27 smurf_paper.tex
update typical bolometer plots + section again
The s8b data post-upgrades is noisier than the old s8d,
but I worked harder at my bolometer selection to illustrate
the features that I was interested in.

\begin{figure*}
\centering
\includegraphics[width=0.49\linewidth]{bolos_point_mix_s2sro.pdf}
\includegraphics[width=0.49\linewidth]{bolos_point_mix.pdf}
\caption{A comparison between single bolometer time series in each of
  the 450 and 850\,\micron\ bands with the mixing chamber temperature
  and azimuth/elevation pointing offsets, before and after upgrading
  the instrument. The grey signals overplotted in the top two panels
  show the residual time-series after removing the common-mode
  signals.  Left: data taken during the S2SRO period, observation 29
  on 20100313. There is a strong correlation between the bolometers
  and the roughly $\sim$25\,s oscillation in the fridge, but not with
  the telescope motion. Also note that the total power in this primary
  signal is similar at both 450 and 850\,\micron, pointing to an
  internal rather than an external source.  The nearly flat
  common-mode subtracted signals show: (i) that most of the
  low-frequency signal is common to all of the bolometers; and (ii)
  the non-correlated, and predominantly white noise at 450\,\micron\
  is significantly larger than at 850\,\micron. Right: data taken with
  the fully-commissioned instrument, observation 38 on
  20111112. Unlike the S2SRO data, there is no strong signal produced
  by variations in the fridge temperature. The 450\,\micron\ data have
  minimal correlated low-frequency noise (as evidenced by the lack of
  a difference between the raw and common-mode subtracted signals),
  although the 850\,\micron data show a common signal correlated with
  the telescope motion (note that this scan has a significantly larger
  amplitude than the S2SRO data set).}
\label{fig:bolos_mix}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.49\linewidth]{pspec_s2sro.pdf}
\includegraphics[width=0.49\linewidth]{pspec.pdf}
\caption{Bolometer power spectral densities for the same two data sets
  (before and after upgrades) used in Fig.~\ref{fig:bolos_mix}. Four
  representative bolometers have been selected at each wavelength, and
  the flat-fielded (but otherwise raw) PSDs are shown as coloured
  dotted lines (the blue signals are for the same time series as those
  shown in Fig.~\ref{fig:bolos_mix}).  The solid black lines are the
  PSDs of the common-mode signals at each wavelength, and the solid
  coloured lines show the PSDs of the bolometers once the common-mode
  is removed.  Finally, the dashed black lines show the spectral shape
  produced by a point source given the scan speeds for the two
  observations (120\,arcsec\,s$^{-1}$ in the left-hand plot, and
  190\,arcsec\,s$^{-1}$ in the right-hand plot); for reference, the
  top horizontal axes shows the conversion from frequency to angular
  scale. Horizontal dotted lines at $2 \times 10^5$ and $5 \times
  10^4$\,pW\,Hz$^{-1}$ at 450 and 850\,\micron, respectively, are
  provided as a visual reference for the white-noise levels. In
  addition to $1/f$ and white noise components, and line features, all
  of the PSDs exhibit the gradual roll-off of the anti-aliasing filter
  just below the Nyquist frequency.  Left: for the S2SRO data there
  are broad line features in the PSDs at both wavelengths above
  $\sim$35\,Hz.  At lower frequencies, the bolometer signals exhibit
  clear $1/f$ knees at approximately 2 and 4\,Hz at 450 and
  850\,\micron. Common-mode subtraction removes most of the correlated
  fridge oscillation signal, lowering the $1/f$ knees to approximately
  0.4 and 2\,Hz (or angular scales of 1.7 and 8,3\,arcmin) at 450 and
  850\,\micron, respectively. Right: data from the fully-commissioned
  instrument tend to have lower $1/f$ knees, and fewer line
  features. The remaining low-frequency noise, now that the fridge
  oscillations have been removed, is considerably more independent
  from bolometer to bolometer than in the S2SRO data (note the larger
  spread in the dotted lines at low frequencies compared to the
  left-hand plot). Finally, this example shows how much the noise in
  the arrays can vary over time. Despite selecting bolometers from the
  same subarrays in each example, these particular 850\,\micron\ data
  from the fully-commissioned instrument are somewhat noisier.}
\label{fig:pspec}
\end{figure*}

@@

In Fig.~\ref{fig:bolos_mix} we show sample time series from single
bolometers in each of the 450 and 850\,micron\ focal planes, as well
as variations in the mixing chamber temperature (as a proxy for the
focal-plane temperature), and the telescope pointing, for two data
sets, before and after the upgrades that followed S2SRO.

In the S2SRO data (left panel), observation 29 on 20100313, both
bolometers share significant long-timescale structure ($\gsim10$\,s)
that appears to be related to variations in the fridge base
temperature, although the similarity is clearly greater at
850\,\micron. In this particular case, the total power in the
fluctuations at 450\,\micron, from $-0.04$\,pW to $+0.08$\,pW, are
larger than the $-0.03$\,pW to $+0.05$\,pW fluctuations at
850\,\micron. Such behaviour might be expected if there is a
comparable varying thermal load from the fridge at each wavelength,
but a larger contribution of atmospheric variations through the
450\,\micron\ bandpass filters. We also note that there is no obvious
strong correlation between the low-frequency signal structure, at
either wavelength, with the telescope motion.

@@

Data from the fully commissioned instrument, observation 38 on
20111112, are shown in the right panel of
Fig.~\ref{fig:bolos_mix}. Having solved the fridge oscillation
problem, these data no longer exhibit a correlation with the mixing
chamber (note that variations are still seen in the mixing chamber; it
is likely that these are not observed at the focal plane itself, where
the NTD thermometry used in the servo loop is performed). There is
minimal correlated low-frequency noise in these 450\,\micron\ data
(note that there is almost no difference once the common-mode has been
subtarcted). The 850\,\micron\ channel, however, exhibits a
significant signal that is obviously correlated with the telescope
motion. This signal is correlated amongs many of the detectors, and
common-mode removal yields fairly flat residuals. Part of the reason
that this is seen here, and not in the S2SRO data set shown, is that
the amplitude of the scan pattern is somewhat larger. Most of this
``scan-synchronous'' noise is attributed to magnetic field pickup, as
described in Section~\ref{sec:magpickup}.

Next, we produce power spectral density (PSD) plots for several
representative bolometers from both focal planes, using the same two
data sets, in Fig.~\ref{fig:pspec}. To produce this figure, we follow
the convention that the PSD as a function of frequency,
$\mathbf{P}(f)$, is normalized such that the integral over frequency
gives the same variance as the time-series variance across the full
time series. In other words, given a bolometer signal $\mathbf{b}(t)$,
%
\begin{equation}
\label{eq:psd}
\langle\mathbf{b}^2(t)\rangle = 2 \int_0^{f_\mathrm{N}} \mathbf{P}(f)
df ,
\end{equation}
%
where we only integrate over the positive frequencies up to
$f_\mathrm{N}$, the Nyquist frequency, and the factor of 2 accounts
for the (equal) power that appears at negative frequencies in the
discrete Fourier Transform. The units of the PSD written in this form
are pW$^2$\,Hz$^{-1}$; without the factor of 2 in front the convention
would be to write the units as pW$^2$\,s to distinguish the two
normalisations.

The dotted coloured lines in Fig~\ref{fig:pspec} show the PSDs for
raw, though flat-fielded data. At each wavelength, and in both data
sets, there are clear $1/f$ knees at frequencies ranging from roughly
1 to 10\,Hz, followed by a predominantly white spectrum punctuated by
line features at higher frequencies, and finally a roll-off caused by
the anti-aliasing filter above $\gsim 70$\,Hz. As indicated in the
previous section, the correlation between the low-frequency components
of the different bolometer signals is large. The solid black lines in
Fig.~\ref{fig:pspec} indicate the PSDs of the common-modes
$\mathbf{c}(t)$ at each wavelength, which reproduce much of the
low-frequency structure, as well as some of the higher-frequency line
features. The $\mathbf{c}(t)$ otherwise drop substantially below the
individual bolometer PSDs at high-frequency, as expected if the
bolometers are dominated by un-correlated white noise. The common-mode
signals are fit to each bolometer time series and removed as in
Fig.~\ref{fig:bolos_mix}, and the resulting PSDs are shown as solid
coloured lines.  For reference, the top horizontal axes have been
converted to angular scale assuming the scan speeds of each
observation (120\,arcsec\,s$^{-1}$ in the left-hand plot, and
190\,arcsec\,s$^{-1}$ in the right-hand plot). The power spectra of
unresolved point-sources are also shown as dashed lines in each band
(arbitrarily normalized), showing that the smallest features
resolvable by the telescope are only minimally affected by the excess
noise in the line features.

@@

Raw data from the fully-commissioned instrument (right-hand panel of
Fig.~\ref{fig:pspec}) generally have noticeably lower $1/f$ knees
(with the exception of some of the noisier 850\,\micron\ bolometers in
this example) as compared with the S2SRO data. The residual
low-frequency noise is considerably less correlated amongst detectors
(larger spread in the dotted colour lines), leading to a less drastic
improvement upon common-mode removal. However, since the data are less
dominated by low-frequency noise to begin with, the signals generally
require less low-frequency filtering to produce maps, and therefore
retain larger-scale structures than with the S2SRO data. Furthermore,
there are generally fewer line features in the PSDs of bolometers in
the fully-commissioned instrument. Finally, these two data sets
illustrate how the white noise performance (NEFD) can vary. For this
particular data set, the 850\,\micron\ noise performance is
substantially worse than in the example from the S2SRO
period. Generally speaking such noise steps occur in only a subset of
the \scuba\ subarrays. However, given the 4-fold increase in available
working bolometers since S2SRO, and the improved low-frequency noise
performance, \scuba\ is now a significantly faster mapping instrument,
and preserves information up to larger angular scales.

>>>>

\begin{figure*}
\includegraphics[width=0.49\linewidth]{bolos_point_mix_s2sro.pdf}
\includegraphics[width=0.49\linewidth]{bolos_point_mix.pdf}
\caption{A comparison between single bolometer time series in each of
  the 450 and 850\,\micron\ bands with the mixing chamber temperature
  and azimuth/elevation pointing offsets, before and after upgrading
  the instrument. The grey signals overplotted in the top two panels
  show the residual time-series after removing the common-mode
  signals.  Left: data taken during the S2SRO period, observation 29
  on 20100313. There is a strong correlation between the bolometers
  and the roughly $\sim$25\,s oscillation in the fridge, but only a
  minor correlation with the telescope motion. The nearly flat
  common-mode subtracted signals show: (i) that most of the
  low-frequency signal is common to all of the bolometers; and (ii)
  the non-correlated, and predominantly white noise at 450\,\micron\
  is significantly larger than at 850\,\micron. Right: data taken with
  the fully-commissioned instrument, observation 38 on
  20111112. Unlike the S2SRO data, there is no strong signal produced
  by variations in the fridge temperature. The 450\,\micron\ data have
  minimal correlated low-frequency noise (as evidenced by the lack of
  a difference between the raw and common-mode subtracted signals),
  although the 850\,\micron\ data show a common signal correlated with
  the telescope motion (note that this scan has a significantly larger
  amplitude than the S2SRO data set), mostly likely caused by magnetic
  field pickup (see Section~\ref{sec:magpickup} and
  Fig.~\ref{fig:magpickup}).}
\label{fig:bolos_mix}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.49\linewidth]{pspec_s2sro.pdf}
\includegraphics[width=0.49\linewidth]{pspec.pdf}
\caption{Bolometer power spectral densities for the same two data sets
  (before and after upgrades) used in Fig.~\ref{fig:bolos_mix}. Four
  of the most sensitive bolometers have been selected at each
  wavelength, and the flat-fielded (but otherwise raw) PSDs are shown
  as coloured dotted lines (the blue signals are for the same time
  series as those shown in Fig.~\ref{fig:bolos_mix}).  The solid black
  lines are the PSDs of the common-mode signals at each wavelength,
  and the solid coloured lines show the PSDs of the bolometers once
  the common-mode is removed.  Finally, the dashed black lines show
  the spectral shape produced by a point source given the scan speeds
  for the two observations (120\,arcsec\,s$^{-1}$ in the left-hand
  plot, and 190\,arcsec\,s$^{-1}$ in the right-hand plot); for
  reference, the top horizontal axes shows the conversion from
  frequency to angular scale. Horizontal dotted lines at $2 \times
  10^5$ and $1 \times 10^5$\,pW\,Hz$^{-1}$ at 450 and 850\,\micron,
  respectively, are provided as a visual reference for the white-noise
  levels. In addition to $1/f$ and white noise components, and line
  features, all of the PSDs exhibit the gradual roll-off of the
  anti-aliasing filter just below the Nyquist frequency.  Left: for
  the S2SRO data there are broad line features in the PSDs at both
  wavelengths above $\sim$35\,Hz.  At lower frequencies, the bolometer
  signals exhibit clear $1/f$ knees at approximately 1 and 2\,Hz at
  450 and 850\,\micron. Common-mode subtraction removes most of the
  correlated fridge oscillation signal, lowering the $1/f$ knees to
  approximately 0.2 and 0.7\,Hz at 450 and 850\,\micron,
  respectively. Right: data from the fully-commissioned instrument
  tend to have lower $1/f$ knees, and fewer line features. The
  remaining low-frequency noise, now that the fridge oscillations have
  been removed, is considerably more independent from bolometer to
  bolometer than in the S2SRO data (note the larger spread in the
  dotted lines at low frequencies compared to the left-hand plot,
  particularly at 450\,\micron). Finally, this example shows that the
  white noise performance is in factor similar for the two subarrays
  (s4a and s8b) that were used both before and after the upgrades.}
\label{fig:pspec}
\end{figure*}

@@

In Fig.~\ref{fig:bolos_mix} we show sample time series from single
bolometers in each of the 450 and 850\,\micron\ focal planes, as well
as variations in the mixing chamber temperature (as a proxy for the
focal-plane temperature), and the telescope pointing, for two data
sets, before and after the upgrades that followed S2SRO.

In the S2SRO data (left panel), observation 29 on 20100313, both
bolometers share significant long-timescale structure ($\gsim10$\,s)
that appears to be related to variations in the fridge base
temperature, although the similarity is clearly greater at
850\,\micron. In this particular case, the total power in the
fluctuations at 450\,\micron, are comparable to those at
850\,\micron. Such behaviour might be expected if there is a
comparable varying thermal load from the fridge at each wavelength
that dominates. We also note that there is no obvious strong
correlation between the low-frequency signal structure, at either
wavelength, with the telescope motion.

@@

Data from the fully commissioned instrument, observation 38 on
20111112, are shown in the right panel of
Fig.~\ref{fig:bolos_mix}. Having solved the fridge oscillation
problem, these data no longer exhibit a correlation with the mixing
chamber (note that variations are still seen in the mixing chamber; it
is likely that these are not observed at the focal plane itself, where
the NTD thermometry used in the servo loop is performed). There is
minimal correlated low-frequency noise in these 450\,\micron\ data
(note that there is almost no difference once the common-mode has been
subtarcted). The 850\,\micron\ channel, however, exhibits a more
significant signal that is obviously correlated with the telescope
motion. It is also correlated amongs many of the detectors, and
common-mode removal corrects it to some extent. Part of the reason
that this is seen here, and not in the S2SRO data set shown, is that
the amplitude of the scan pattern is larger. Most of this
``scan-synchronous'' noise is attributed to magnetic field pickup, as
described in Section~\ref{sec:magpickup}.

Next, we produce power spectral density (PSD) plots for four of the
most sensitive bolometers from both focal planes, using the same two
data sets, in Fig.~\ref{fig:pspec}. To produce this figure, we follow
the convention that the PSD as a function of frequency,
$\mathbf{P}(f)$, is normalized such that the integral over frequency
gives the same variance as the time-series variance across the full
time series. In other words, given a bolometer signal $\mathbf{b}(t)$,
%
\begin{equation}
\label{eq:psd}
\langle\mathbf{b}^2(t)\rangle = 2 \int_0^{f_\mathrm{N}} \mathbf{P}(f)
df ,
\end{equation}
%
where we only integrate over the positive frequencies up to
$f_\mathrm{N}$, the Nyquist frequency, and the factor of 2 accounts
for the (equal) power that appears at negative frequencies in the
discrete Fourier Transform. The units of the PSD written in this form
are pW$^2$\,Hz$^{-1}$; without the factor of 2 in front the convention
would be to write the units as pW$^2$\,s to distinguish the two
normalisations.

The dotted coloured lines in Fig~\ref{fig:pspec} show the PSDs for
raw, though flat-fielded data. At each wavelength, and in both data
sets, there are clear $1/f$ knees at frequencies ranging from roughly
1 to 2\,Hz, followed by a predominantly white spectrum punctuated by
line features at higher frequencies, and finally a roll-off caused by
the anti-aliasing filter above $\gsim 70$\,Hz. As indicated in the
previous section, the correlation between the low-frequency components
of the different bolometer signals is large. The solid black lines in
Fig.~\ref{fig:pspec} indicate the PSDs of the common-modes
$\mathbf{c}(t)$ at each wavelength, which reproduce much of the
low-frequency structure, as well as some of the higher-frequency line
features. The $\mathbf{c}(t)$ otherwise drop substantially below the
individual bolometer PSDs at high-frequency, as expected if the
bolometers are dominated by un-correlated white noise in that part of
the spectrum. The common-mode signals are fit to each bolometer time
series and removed as in Fig.~\ref{fig:bolos_mix}, and the resulting
PSDs are shown as solid coloured lines.  For reference, the top
horizontal axes have been converted to angular scale assuming the scan
speeds of each observation (120\,arcsec\,s$^{-1}$ in the left-hand
plot, and 190\,arcsec\,s$^{-1}$ in the right-hand plot). The power
spectra of unresolved point-sources are also shown as dashed lines in
each band (arbitrarily normalized), showing that the smallest features
resolvable by the telescope are only minimally affected by the excess
noise in the line features at these scan speeds (this may not be the
case at higher scan speeds).

@@

Raw data from the fully-commissioned instrument (right-hand panel of
Fig.~\ref{fig:pspec}) generally have less significant $1/f$ noise as
compared with the S2SRO data. The residual low-frequency noise is
considerably less correlated amongst detectors (larger spread in the
dotted colour lines, particularly noticeable in these 450\,\micron\
data), leading to a less drastic improvement upon common-mode
removal. However, since the data are less dominated by low-frequency
noise to begin with, the signals generally require less low-frequency
filtering to produce maps, and therefore retain larger-scale
structures than with the S2SRO data. Furthermore, there are generally
fewer line features in the PSDs of bolometers in the
fully-commissioned instrument. Finally, these two data sets illustrate
that the white noise performance (NEFD) is in fact similar before and
the upgrades for these two subarrays (s4a at 450\,micron, and s8b at
850\,\micron). The main improvements are reduction in correlated and
line noise features mentioned above, and the larger number of working
bolometers (approximately a factor of 4).
